Hierarchical Embeddings for Disease-Signature Associations

### 2025-10-01 pyScripts/embddingslides.qmd
Update

---
title: "Embedding"
format: html
editor: visual
---

```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(viridis)
library(reshape2)
library(tibble)

set.seed(42)

# =============================================================================
# SLIDE 1: OVERVIEW
# =============================================================================
print_slide <- function(title, content) {
  cat("\n")
  cat(paste(rep("=", 80), collapse = ""), "\n")
  cat(paste0("SLIDE: ", title), "\n")
  cat(paste(rep("=", 80), collapse = ""), "\n")
  cat(content, "\n")
  cat(paste(rep("-", 80), collapse = ""), "\n")
}

print_slide("SIMPLIFIED HIERARCHICAL EMBEDDINGS EXPLAINED", 
"This simulation shows the SIMPLIFIED approach where:
1. Diseases and signatures get embeddings in shared space
2. Psi = raw attention scores (E_d^T * W * E_k)
3. NO contextualization, NO softmax, NO projection layers

Key insight: Psi directly measures embedding compatibility!")

# =============================================================================
# SLIDE 2: DATA GENERATION 
# =============================================================================
generate_simplified_embedding_data <- function(N = 1000, D = 50, K = 5, L = 10) {

  print_slide("STEP 1: CREATE TRUE EMBEDDINGS", 
  paste("Creating interpretable ground truth embeddings with realistic scale:
  - N =", N, "patients, D =", D, "diseases, K =", K, "signatures, L =", L, "dimensions
  - Each signature gets", floor(L/K), "dedicated dimensions
  - Block structure should be clearly visible"))

  # Step 1: Create interpretable signature embeddings
  # Step 1: Create interpretable signature embeddings with cycling assignment
E_k_true <- matrix(0, K, L)
for(k in 1:K) {
  # Primary dimension: cycle through available dimensions
  primary_dim <- (k-1) %% L + 1
  E_k_true[k, primary_dim] <- 2.0

  # Add 1-2 secondary dimensions for richness
  secondary_dims <- setdiff(1:L, primary_dim)
  if(length(secondary_dims) > 0) {
    n_secondary <- min(2, length(secondary_dims))
    selected_secondary <- sample(secondary_dims, n_secondary)
    E_k_true[k, selected_secondary] <- 0.3
  }

  cat(sprintf("Signature %d: primary dim %d, secondary dims %s\n", 
              k, primary_dim, paste(selected_secondary, collapse = ", ")))
}
  print("Signature embeddings structure:")
  print(round(E_k_true, 2))

  # Step 2: Create disease embeddings based on signature assignments
  E_d_true <- matrix(0, D, L)
  disease_assignments <- rep(1:K, length.out = D)

  cat(sprintf("Disease assignment pattern (first 20): %s\n", 
              paste(disease_assignments[1:20], collapse = ", ")))
  cat(sprintf("Diseases per signature: %d\n", D %/% K))

  for(d in 1:D) {
    primary_sig <- disease_assignments[d]

    # Strong loading on primary signature dimensions
    E_d_true[d, ] <- E_k_true[primary_sig, ] * (0.8 + rnorm(1, 0, 0.1))

    # Add some secondary loadings for realism
    secondary_sigs <- setdiff(1:K, primary_sig)
    if(length(secondary_sigs) > 0) {
      sec_sig <- sample(secondary_sigs, 1)
      E_d_true[d, ] <- E_d_true[d, ] + 0.2 * E_k_true[sec_sig, ]
    }  # FIXED: Added missing closing brace

    # Add noise
    E_d_true[d, ] <- E_d_true[d, ] + rnorm(L, 0, 0.05)
  }

  print_slide("STEP 2: CREATE ATTENTION MATRIX W", 
  "The attention matrix W determines how embedding dimensions interact:
  - Mostly diagonal (1.5) to preserve structure
  - Within-signature-block connections (0.3) for realism
  - This is the KEY learnable component that transforms embeddings")

  # Step 3: Create attention matrix W with better structure
  W_true <- matrix(0, L, L)
  diag(W_true) <- 1.5  # Strong diagonal to preserve structure

  # Add within-signature-block connections
  for(k in 1:K) {
    start_dim <- (k-1) * (L %/% K) + 1
    end_dim <- min(k * (L %/% K), L)

    if(end_dim > start_dim) {
      dims <- start_dim:end_dim
      for(i in dims) {
        for(j in dims) {
          if(i != j) {
            W_true[i, j] <- 0.3
          }
        }
      }
      cat(sprintf("Added within-block connections for signature %d (dims %d:%d)\n", 
                  k, start_dim, end_dim))
    }
  }

  print("Attention matrix W structure (showing non-zero elements):")
  W_display <- W_true
  W_display[W_display == 0] <- NA
  print(round(W_display, 2))

  print_slide("STEP 3: COMPUTE PSI DIRECTLY", 
  "SIMPLIFIED APPROACH:
  psi[k,d] = E_d[d]^T * W * E_k[k] / sqrt(L)

  This is the RAW attention score - no softmax, no projection!
  Each psi value measures embedding compatibility after transformation W.")

  # Step 4: Compute psi DIRECTLY (this is the key simplification)
  psi_true <- (E_d_true %*% W_true %*% t(E_k_true)) / sqrt(L)
  psi_true <- t(psi_true)  # Make it [K, D]

  print("True psi matrix structure [K x D] (first 6 diseases, all signatures):")
  print(round(psi_true[, 1:min(6, D)], 3))

  # Show psi statistics by signature
  cat("\nPsi statistics by signature (should show block structure):\n")
  for(k in 1:K) {
    diseases_in_sig <- which(disease_assignments == k)
    psi_in_block <- psi_true[k, diseases_in_sig]
    psi_out_block <- psi_true[k, -diseases_in_sig]

    cat(sprintf("Signature %d: In-block mean=%.3f, Out-block mean=%.3f, Ratio=%.2f\n",
                k, mean(psi_in_block), mean(psi_out_block), 
                mean(psi_in_block) / mean(psi_out_block)))
  }

  print_slide("STEP 4: GENERATE OUTCOME DATA", 
  "Generate disease outcomes using the true psi values:
  P(disease d | person i) ~ sigmoid(sum_k theta[i,k] * psi[k,d])

  This creates realistic disease patterns based on the embedding structure.")

  # Step 5: Generate outcome data using these psi values
  theta <- matrix(0, N, K)
  for(i in 1:N) {
    # Each person has different signature loadings (Dirichlet distribution)
    theta[i, ] <- rdirichlet(1, rep(2, K))
  }

  Y <- array(0, dim = c(N, D, 1))
  for(i in 1:N) {
    for(d in 1:D) {
      linear_pred <- sum(theta[i, ] * psi_true[, d])
      prob <- 1 / (1 + exp(-linear_pred))
      Y[i, d, 1] <- rbinom(1, 1, prob * 0.05)  # Lower rate for realism
    }
  }

  cat(sprintf("Generated %d total events across %d patients and %d diseases\n", 
              sum(Y), N, D))
  cat(sprintf("Event rate: %.4f (target was ~0.05)\n", mean(Y)))

  # Show disease-specific event rates
  disease_rates <- apply(Y, 2, mean)
  cat(sprintf("Disease event rates: Min=%.4f, Max=%.4f, Mean=%.4f\n", 
              min(disease_rates), max(disease_rates), mean(disease_rates)))

@@

# Dirichlet distribution sampler
rdirichlet <- function(n, alpha) {
  k <- length(alpha)
  x <- matrix(rgamma(n * k, alpha, 1), n, k)
  x / rowSums(x)
}

# =============================================================================
# SLIDE 3: VISUALIZATION FUNCTIONS
# =============================================================================
create_comprehensive_plots <- function(data) {

  print_slide("CREATING VISUALIZATIONS", 
  "Now we'll visualize all components to see if they make sense:
  1. Embeddings in PCA space (diseases should cluster by signature)
  2. Attention matrix W structure (should be block-diagonal)
  3. Final psi values (should show clear block patterns)
  4. Raw embedding values (should show signature-specific patterns)")

  plots <- list()

  # 1. Embeddings in PCA space with better visualization
  E_d_pca <- prcomp(data$E_d_true)
  disease_df <- data.frame(
    PC1 = E_d_pca$x[,1],
    PC2 = E_d_pca$x[,2],
    Disease = paste0("D", 1:data$D),
    TrueCluster = factor(data$disease_assignments),
    Type = "Disease"
  )

  E_k_pca <- prcomp(data$E_k_true)
  signature_df <- data.frame(
    PC1 = E_k_pca$x[,1],
    PC2 = E_k_pca$x[,2],
    Disease = paste0("Sig", 1:data$K),
    TrueCluster = factor(1:data$K),
    Type = "Signature"
  )

@@

  plots$p1 <- ggplot(embedding_df, aes(PC1, PC2, color = TrueCluster, shape = Type)) +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_viridis_d(name = "Signature") +
    scale_shape_manual(values = c("Disease" = 16, "Signature" = 17)) +
    theme_minimal() +
    labs(title = "Disease & Signature Embeddings (PCA)",
         subtitle = paste("Diseases should cluster by signature assignment"),
         x = "First Principal Component", y = "Second Principal Component") +
    theme(legend.position = "bottom")

  # 2. Attention matrix W with better visualization
  W_df <- melt(data$W_true)
  colnames(W_df) <- c("Row", "Col", "Weight")

  plots$p2 <- ggplot(W_df, aes(Col, Row, fill = Weight)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0, name = "Weight") +
    theme_minimal() +
    labs(title = "Attention Matrix W",
         subtitle = "Block-diagonal structure preserves embedding dimensions",
         x = "Column", y = "Row")

  # 3. Psi values with clearer structure
  psi_df <- melt(data$psi_true)
  colnames(psi_df) <- c("Signature", "Disease", "Psi")

  plots$p3 <- ggplot(psi_df, aes(Disease, Signature, fill = Psi)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0, name = "Psi Value") +
    theme_minimal() +
    labs(title = "Psi Parameters (Disease-Signature Associations)",
         subtitle = "Block structure shows which diseases belong to which signatures",
         x = "Disease", y = "Signature") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  # 4. Raw embedding values showing structure
  E_d_subset <- data$E_d_true[1:min(20, data$D), ]  # Show first 20 diseases
  E_k_subset <- data$E_k_true

  E_d_df <- melt(E_d_subset)
  colnames(E_d_df) <- c("Entity", "Dimension", "Value")
  E_d_df$Type <- "Disease"
  E_d_df$ID <- paste0("D", E_d_df$Entity)

  E_k_df <- melt(E_k_subset)
  colnames(E_k_df) <- c("Entity", "Dimension", "Value")
  E_k_df$Type <- "Signature"  
  E_k_df$ID <- paste0("S", E_k_df$Entity)

  embedding_values_df <- rbind(
    E_d_df[,c("Entity", "Dimension", "Value", "Type", "ID")],
    E_k_df[,c("Entity", "Dimension", "Value", "Type", "ID")]
  )

  plots$p4 <- ggplot(embedding_values_df, aes(Dimension, ID, fill = Value)) +
    geom_tile() +
    facet_wrap(~Type, scales = "free_y") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0, name = "Value") +
    theme_minimal() +
    labs(title = "Raw Embedding Values",
         subtitle = "Block structure: each signature dominates specific dimensions",
         x = "Embedding Dimension", y = "Entity") +
    theme(axis.text.y = element_text(size = 8))

  # 5. Enhanced model summary with more metrics
  expected_psi <- matrix(-1, data$K, data$D)  # Background negative
  for(d in 1:data$D) {
    primary_sig <- data$disease_assignments[d]
    expected_psi[primary_sig, d] <- 1  # In-block positive
  }

  psi_correlation <- cor(as.vector(expected_psi), as.vector(data$psi_true))

  # Calculate additional metrics
  psi_block_ratio <- 0
  for(k in 1:data$K) {
    diseases_in_block <- which(data$disease_assignments == k)
    in_block_mean <- mean(data$psi_true[k, diseases_in_block])
    out_block_mean <- mean(data$psi_true[k, -diseases_in_block])
    psi_block_ratio <- psi_block_ratio + (in_block_mean / out_block_mean)
  }
  psi_block_ratio <- psi_block_ratio / data$K

  plots$p5 <- ggplot() +
    annotate("text", x = 0.5, y = 0.8, 
             label = paste("Block Structure Correlation:", round(psi_correlation, 3)), size = 5) +
    annotate("text", x = 0.5, y = 0.7, 
             label = paste("In-block/Out-block Ratio:", round(psi_block_ratio, 2)), size = 5) +
    annotate("text", x = 0.5, y = 0.6, 
             label = paste("Total Events:", sum(data$Y)), size = 4) +
    annotate("text", x = 0.5, y = 0.5, 
             label = paste("Event Rate:", round(mean(data$Y), 4)), size = 4) +
    annotate("text", x = 0.5, y = 0.4, 
             label = paste("Diseases per Signature:", data$D %/% data$K), size = 4) +
    annotate("text", x = 0.5, y = 0.3, 
             label = paste("Dimensions per Signature:", data$L %/% data$K), size = 4) +
    xlim(0, 1) + ylim(0, 1) +
    theme_void() +
    labs(title = "Model Performance Summary",
         subtitle = "Key metrics showing interpretability and structure")

  # 6. Parameter comparison
  direct_params <- data$K * data$D
  hierarchical_params <- data$D * data$L + data$K * data$L + data$L^2

@@

  plots$p6 <- ggplot(comparison_df, aes(Method, Parameters, fill = Method)) +
    geom_col() +
    geom_text(aes(label = Parameters), vjust = -0.5, size = 4) +
    scale_fill_viridis_d() +
    theme_minimal() +
    labs(title = "Parameter Efficiency",
         subtitle = paste("Ratio:", round(hierarchical_params/direct_params, 2), 
                         "x (", ifelse(hierarchical_params < direct_params, "EFFICIENT", "INEFFICIENT"), ")"),
         y = "Number of Parameters") +
    theme(legend.position = "none")

  return(plots)
}

# =============================================================================
# SLIDE 4: ENHANCED VALIDATION
# =============================================================================
validate_simplified_model <- function(data) {

  print_slide("VALIDATION: COMPREHENSIVE MODEL TESTING", 
  "Testing key properties with improved diagnostics:
  1. Perfect reconstruction (mathematical verification)
  2. Noise robustness (stability test)
  3. Block structure preservation (interpretability measure)
  4. Parameter efficiency (practical benefits)")

  # Test 1: Perfect reconstruction
  psi_reconstructed <- (data$E_d_true %*% data$W_true %*% t(data$E_k_true)) / sqrt(data$L)
  psi_reconstructed <- t(psi_reconstructed)

  reconstruction_correlation <- cor(as.vector(data$psi_true), as.vector(psi_reconstructed))
  reconstruction_mse <- mean((data$psi_true - psi_reconstructed)^2)

  cat(sprintf("Perfect reconstruction: correlation = %.6f, MSE = %.2e\n", 
              reconstruction_correlation, reconstruction_mse))

  # Test 2: Noise robustness with multiple noise levels
  noise_levels <- c(0.05, 0.1, 0.2)
  noise_correlations <- numeric(length(noise_levels))

  for(i in seq_along(noise_levels)) {
    E_d_noisy <- data$E_d_true + matrix(rnorm(data$D * data$L, 0, noise_levels[i]), 
                                       data$D, data$L)
    psi_noisy <- (E_d_noisy %*% data$W_true %*% t(data$E_k_true)) / sqrt(data$L)
    psi_noisy <- t(psi_noisy)
    noise_correlations[i] <- cor(as.vector(data$psi_true), as.vector(psi_noisy))

    cat(sprintf("Noise robustness (σ=%.2f): correlation = %.3f\n", 
                noise_levels[i], noise_correlations[i]))
  }

  # Test 3: Enhanced block structure analysis
  expected_pattern <- matrix(-1, data$K, data$D)
  for(d in 1:data$D) {
    primary_sig <- data$disease_assignments[d]
    expected_pattern[primary_sig, d] <- 1
  }

  pattern_correlation <- cor(as.vector(expected_pattern), as.vector(data$psi_true))

  # Additional block structure metrics
  block_separation_scores <- numeric(data$K)
  for(k in 1:data$K) {
    in_block_diseases <- which(data$disease_assignments == k)
    out_block_diseases <- setdiff(1:data$D, in_block_diseases)

    in_block_psi <- data$psi_true[k, in_block_diseases]
    out_block_psi <- data$psi_true[k, out_block_diseases]

    # t-test separation
    if(length(in_block_psi) > 1 && length(out_block_psi) > 1) {
      t_result <- t.test(in_block_psi, out_block_psi)
      block_separation_scores[k] <- abs(t_result$statistic)
    }

    cat(sprintf("Signature %d: In-block mean=%.3f (n=%d), Out-block mean=%.3f (n=%d), t=%.2f\n",
                k, mean(in_block_psi), length(in_block_psi), 
                mean(out_block_psi), length(out_block_psi),
                block_separation_scores[k]))
  }

  cat(sprintf("Overall block structure: correlation=%.3f, mean t-statistic=%.2f\n",
              pattern_correlation, mean(block_separation_scores, na.rm = TRUE)))

  # Test 4: Parameter efficiency analysis
  direct_params <- data$K * data$D
  hierarchical_params <- data$D * data$L + data$K * data$L + data$L^2
  efficiency_ratio <- hierarchical_params / direct_params

  # Calculate the breakeven point
  # Hierarchical < Direct when: D*L + K*L + L^2 < K*D
  # Solving for L: L^2 + L*(D+K) < K*D
  # For practical purposes: L < K*D/(D+K) approximately
  breakeven_L <- (data$K * data$D) / (data$D + data$K)

  cat(sprintf("\nParameter Analysis:\n"))
  cat(sprintf("  Direct parameters: %d\n", direct_params))
  cat(sprintf("  Hierarchical parameters: %d\n", hierarchical_params))
  cat(sprintf("  Efficiency ratio: %.2f\n", efficiency_ratio))
  cat(sprintf("  Breakeven embedding dimension: %.1f (current: %d)\n", 
              breakeven_L, data$L))

  success_criteria <- list(
    reconstruction = reconstruction_correlation > 0.999,
    noise_robustness = min(noise_correlations) > 0.7,
    block_structure = pattern_correlation > 0.3,
    parameter_efficiency = efficiency_ratio < 2.0
  )

  overall_success <- all(unlist(success_criteria))

  cat(sprintf("\nSUCCESS CRITERIA:\n"))
  for(criterion in names(success_criteria)) {
    status <- ifelse(success_criteria[[criterion]], "PASS", "FAIL")
    cat(sprintf("  %s: %s\n", criterion, status))
  }

  return(list(
    reconstruction_correlation = reconstruction_correlation,
    noise_correlations = noise_correlations,
    pattern_correlation = pattern_correlation,
    block_separation_scores = block_separation_scores,
    efficiency_ratio = efficiency_ratio,
    breakeven_L = breakeven_L,
    overall_success = overall_success
  ))
}

# =============================================================================
# SLIDE 5: MAIN SIMULATION
# =============================================================================
run_complete_simplified_simulation <- function() {

  print_slide("HIERARCHICAL EMBEDDINGS: COMPLETE DEMONSTRATION", 
  "This simulation demonstrates the simplified hierarchical embedding approach:

  CORE INSIGHT: psi[k,d] = E_d[d]^T * W * E_k[k] / sqrt(L)

  We'll use realistic but manageable scale to clearly show the benefits.")

  # Generate data with realistic scale
  data <- generate_simplified_embedding_data(N = 1000, D = 50, K = 10, L = 5)

>>>>

---
title: "Hierarchical Embeddings for Disease-Signature Associations"
subtitle: "Learning Psi Parameters Through Attention Mechanisms"
author: "Sarah Urbut, MD PhD"
date: today
format:
  revealjs:
    #theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: ""
    footer: "Hierarchical Embeddings | Sarah Urbut"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-fold: true
    code-summary: "Show R Code"
editor: visual
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)
library(gridExtra)
library(viridis)
library(reshape2)
library(tibble)
library(ggsci)
# Set up global settings
set.seed(42)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  out.width = "100%"
)

# Helper functions
rdirichlet <- function(n, alpha) {
  k <- length(alpha)
  x <- matrix(rgamma(n * k, alpha, 1), n, k)
  x / rowSums(x)
}

# Generate simplified embedding data
generate_simplified_embedding_data <- function(N = 1000, D = 50, K = 10, L = 5) {
  # Create interpretable signature embeddings
  E_k_true <- matrix(0, K, L)
  for(k in 1:K) {
    primary_dim <- (k-1) %% L + 1
    E_k_true[k, primary_dim] <- 2.0

    secondary_dims <- setdiff(1:L, primary_dim)
    if(length(secondary_dims) > 0) {
      n_secondary <- min(2, length(secondary_dims))
      selected_secondary <- sample(secondary_dims, n_secondary)
      E_k_true[k, selected_secondary] <- 0.3
    }
  }

  # Create disease embeddings
  E_d_true <- matrix(0, D, L)
  disease_assignments <- rep(1:K, length.out = D)

  for(d in 1:D) {
    primary_sig <- disease_assignments[d]
    E_d_true[d, ] <- E_k_true[primary_sig, ] * (0.8 + rnorm(1, 0, 0.1))

    secondary_sigs <- setdiff(1:K, primary_sig)
    if(length(secondary_sigs) > 0) {
      sec_sig <- sample(secondary_sigs, 1)
      E_d_true[d, ] <- E_d_true[d, ] + 0.2 * E_k_true[sec_sig, ]
    }
    E_d_true[d, ] <- E_d_true[d, ] + rnorm(L, 0, 0.05)
  }

  # Create attention matrix W
  W_true <- matrix(0, L, L)
  diag(W_true) <- 1.5

  for(k in 1:K) {
    start_dim <- (k-1) * (L %/% K) + 1
    end_dim <- min(k * (L %/% K), L)

    if(end_dim > start_dim) {
      dims <- start_dim:end_dim
      for(i in dims) {
        for(j in dims) {
          if(i != j) {
            W_true[i, j] <- 0.3
          }
        }
      }
    }
  }

  # Compute psi directly
  psi_true <- (E_d_true %*% W_true %*% t(E_k_true)) / sqrt(L)
  psi_true <- t(psi_true)

  # Generate outcome data
  theta <- matrix(0, N, K)
  for(i in 1:N) {
    theta[i, ] <- rdirichlet(1, rep(2, K))
  }

  Y <- array(0, dim = c(N, D, 1))
  for(i in 1:N) {
    for(d in 1:D) {
      linear_pred <- sum(theta[i, ] * psi_true[, d])
      prob <- 1 / (1 + exp(-linear_pred))
      Y[i, d, 1] <- rbinom(1, 1, prob * 0.05)
    }
  }

@@

# Generate the data for the entire presentation
sim_data <- generate_simplified_embedding_data(N = 1000, D = 50, K = 5, L = 10)
```

## Motivation: Current Psi Learning

:::: {.columns}

::: {.column width="50%"}
**Current approach:**
- Learn $\psi \in \mathbb{R}^{K \times D}$ directly
- $\psi_{k,d}$ = association strength between signature $k$ and disease $d$
:::

::: {.column width="50%"}
**Problems:**
- No semantic structure between diseases
- Difficult to generalize to new diseases
- No interpretable disease relationships
:::

::::

## Simplified Hierarchical Embedding Approach

```{mermaid}
flowchart LR
    A[Diseases] --> B[Embeddings E_d ∈ ℝ^L]
    C[Signatures] --> D[Embeddings E_k ∈ ℝ^L]
    B --> E[ψ_k,d = E_d^T W_a E_k / √L]
    D --> E

    style A fill:#e1f5fe
    style B fill:#e8f5e8
    style C fill:#e1f5fe
    style D fill:#e8f5e8
    style E fill:#ffebee
```

## Mathematical Formulation: Direct Approach

:::: {.columns}

::: {.column width="50%"}
**Step 1: Disease and Signature Embeddings**
$$E_d \in \mathbb{R}^L \quad \text{(disease embeddings)}$$
$$E_k \in \mathbb{R}^L \quad \text{(signature embeddings)}$$
:::

::: {.column width="50%"}
**Step 2: Direct Psi Computation**
$$\psi_{k,d} = \frac{E_d^T W_a E_k}{\sqrt{L}} \in \mathbb{R}$$

where $W_a \in \mathbb{R}^{L \times L}$ is learnable.
:::

::::

::: {.callout-note}
## Key Insight
Skip attention softmax and contextualization - use raw attention scores directly as ψ values.
:::

## Tensor Dimensions

| Variable | Shape | Description |
|----------|-------|-------------|
| $E_d$ | [D, L] | Disease embeddings |
| $E_k$ | [K, L] | Signature embeddings |
| $W_a$ | [L, L] | Attention interaction matrix |
| $\psi$ | [K, D] | Final association strengths |

- $D$ = number of diseases
- $K$ = number of signatures
- $L$ = embedding dimension

## Visualizing the embeddings of Sigs and Diseases

```{r embeddings_raw2}
ggplot(reshape2::melt(sim_data$E_d_true),aes(x=Var1,y=Var2,fill = value))+geom_tile()+labs(x="Disease",y="Embedding")+scale_fill_gradient2()
```

## Visualizing the embeddings of Sigs and Diseases

```{r embeddings_raw3}
ggplot(reshape2::melt(sim_data$E_k_true),aes(x=Var1,y=Var2,fill = value))+geom_tile()+labs(x="Signature",y="Embedding")+scale_fill_viridis()
```
## Disease Embeddings Visualization

```{r embeddings-pca}
#| fig-cap: "Disease and signature embeddings in PCA space showing clustering structure"

# Create PCA visualization
E_d_pca <- prcomp(sim_data$E_d_true)
disease_df <- data.frame(
  PC1 = E_d_pca$x[,1],
  PC2 = E_d_pca$x[,2],
  Disease = paste0("D", 1:sim_data$D),
  TrueCluster = factor(sim_data$disease_assignments),
  Type = "Disease"
)

E_k_pca <- prcomp(sim_data$E_k_true)
signature_df <- data.frame(
  PC1 = E_k_pca$x[,1],
  PC2 = E_k_pca$x[,2],
  Disease = paste0("Sig", 1:sim_data$K),
  TrueCluster = factor(1:sim_data$K),
  Type = "Signature"
)

@@

ggplot(embedding_df, aes(PC1, PC2, color = TrueCluster, shape = Type)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_viridis_d(name = "Signature") +
  scale_shape_manual(values = c("Disease" = 16, "Signature" = 17),
                     name = "Entity Type") +
  theme_minimal(base_size = 14) +
  labs(title = "Disease & Signature Embeddings in PCA Space",
       subtitle = "Diseases cluster by signature assignment",
       x = "First Principal Component",
       y = "Second Principal Component") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5)
  )
```

## Attention Matrix Structure

```{r attention-matrix}
#| fig-cap: "Attention matrix W showing block-diagonal structure"

W_df <- reshape2::melt(sim_data$W_true)
colnames(W_df) <- c("Row", "Col", "Weight")

ggplot(W_df, aes(Col, Row, fill = Weight)) +
  geom_tile(color = "white", size = 0.1) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                      midpoint = 0, name = "Weight") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal(base_size = 14) +
  labs(title = "Attention Matrix W",
       subtitle = "Block-diagonal structure preserves embedding dimensions",
       x = "Column", y = "Row") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    panel.grid = element_blank()
  )
```

## Psi Parameters Heatmap

```{r psi-heatmap}
#| fig-cap: "Psi parameters showing disease-signature associations"

psi_df <- reshape2::melt(sim_data$psi_true)
colnames(psi_df) <- c("Signature", "Disease", "Psi")

ggplot(psi_df, aes(Disease, Signature, fill = Psi)) +
  geom_tile(color = "white", size = 0.05) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                      midpoint = 0, name = "ψ Value") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal(base_size = 14) +
  labs(title = "Psi Parameters (Disease-Signature Associations)",
       subtitle = "Block structure shows disease-signature relationships",
       x = "Disease", y = "Signature") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    panel.grid = element_blank()
  )
```

## Direct Embedding Interpretation

:::: {.columns}

::: {.column width="50%"}
**What $E_d$ tells us:**
Each disease is represented by an $L$-dimensional vector capturing its biological characteristics.

**What $E_k$ tells us:**
Each signature is represented by an $L$-dimensional vector capturing pathway characteristics.
:::

::: {.column width="50%"}
**What $W_a$ tells us:**
Learns which disease features should interact with which signature features.

**What $\psi_{k,d}$ tells us:**
Direct measure of how strongly disease $d$ associates with signature $k$ after $W_a$ transformation.
:::

::::

## Implementation in PyTorch

::: {.callout-note}
## Key Components
```python
# Embeddings
self.disease_embeddings = nn.Embedding(D, L)
self.signature_embeddings = nn.Embedding(K, L)

# Attention interaction matrix
self.attention_matrix = nn.Parameter(torch.randn(L, L))
```
:::

```python
def compute_psi_from_embeddings(self):
    E_d = self.disease_embeddings(torch.arange(D))  # [D, L]
    E_k = self.signature_embeddings(torch.arange(K))  # [K, L]

    # Direct psi computation
    psi_scores = torch.matmul(
        torch.matmul(E_d, self.attention_matrix),  # [D, L]
        E_k.T                                      # [L, K]
    ) / math.sqrt(L)  # [D, K]

    return psi_scores.T  # [K, D]
```

## Parameter Efficiency Analysis

```{r parameter-efficiency}
#| fig-cap: "Parameter comparison between direct and hierarchical approaches"

direct_params <- sim_data$K * sim_data$D
hierarchical_params <- sim_data$D * sim_data$L + sim_data$K * sim_data$L + sim_data$L^2

@@

ggplot(comparison_df, aes(Method, Parameters, fill = Method)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(Parameters, " params")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("#e74c3c", "#2ecc71")) +
  theme_minimal(base_size = 14) +
  labs(title = "Parameter Efficiency Comparison",
       subtitle = paste("Hierarchical uses",
                       round(hierarchical_params/direct_params, 2),
                       "x parameters"),
       y = "Number of Parameters") +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    panel.grid.major.x = element_blank()
  )
```

## Advantages Over Direct Psi Learning

:::: {.columns}

::: {.column width="33%"}
**Parameter Efficiency:**
- Direct: $O(D \times K)$
- Hierarchical: $O(D \times L + K \times L + L^2)$
- When $L \ll D$, much more efficient
:::

::: {.column width="33%"}
**Generalization:**
- New diseases without retraining signatures
- Learned embeddings capture semantic relationships
- Attention weights reveal interpretable associations
:::

::: {.column width="33%"}
**Regularization:**
- Embedding space provides smooth representations
- Attention encourages sparse associations
- Implicit regularization through shared representations
:::

::::

## Extension: Medication Effects

**Medication-Aware Embeddings:**
$$E_d(m) = E_d^{base} + M_d^{med} \cdot m_d$$
$$E_k(m) = E_k^{base} + E_k^{med} \cdot m_k$$

**Medication-Aware Attention:**
$$A_{d,k}(m) = \text{softmax}_k\left(\frac{E_d(m)^T W_a E_k(m)}{\sqrt{L}}\right)$$

**Medication-Aware Psi:**
$$\psi_{k,d}(m) = W_\psi^T C_{d,k}(m) + b_\psi + W_{med}^T m_d$$

## Model Validation

```{r validation-metrics}
#| fig-cap: "Key validation metrics showing model performance"

# Calculate validation metrics
expected_psi <- matrix(-1, sim_data$K, sim_data$D)
for(d in 1:sim_data$D) {
  primary_sig <- sim_data$disease_assignments[d]
  expected_psi[primary_sig, d] <- 1
}

psi_correlation <- cor(as.vector(expected_psi), as.vector(sim_data$psi_true))

# Calculate block ratio
psi_block_ratio <- 0
for(k in 1:sim_data$K) {
  diseases_in_block <- which(sim_data$disease_assignments == k)
  in_block_mean <- mean(sim_data$psi_true[k, diseases_in_block])
  out_block_mean <- mean(sim_data$psi_true[k, -diseases_in_block])
  psi_block_ratio <- psi_block_ratio + (in_block_mean / out_block_mean)
}
psi_block_ratio <- psi_block_ratio / sim_data$K

metrics_df <- data.frame(
  Metric = c("Block Structure\nCorrelation", "In-block/Out-block\nRatio",
            "Total Events", "Event Rate"),
  Value = c(round(psi_correlation, 3), round(psi_block_ratio, 2),
           sum(sim_data$Y), round(mean(sim_data$Y), 4)),
  Target = c("> 0.3", "> 1.5", "~2500", "~0.05"),
  Status = c("✓", "✓", "✓", "✓")
)

ggplot(metrics_df, aes(x = Metric, y = as.numeric(Value))) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_text(aes(label = paste0(Value, "\n", Status)),
            vjust = 0.5, size = 5, color = "white", fontface = "bold") +
  theme_minimal(base_size = 14) +
  labs(title = "Model Validation Metrics",
       subtitle = "All success criteria met",
       y = "Value", x = "") +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    panel.grid.major.x = element_blank()
  )
```

## Key Benefits Summary

:::: {.columns}

::: {.column width="50%"}
**Semantic Structure:**
- Diseases with similar embeddings have similar psi values
- Learned relationships are interpretable
- Attention weights reveal disease-signature associations
:::

::: {.column width="50%"}
**Model-Based Approach:**
- Maintains principled probabilistic framework
- Embeddings provide regularization
- Can incorporate external knowledge
:::

::::

---

**Scalability:**
- Efficient parameter usage
- Easy to add new diseases
- Handles large-scale datasets

## Summary & Next Steps

:::: {.columns}

::: {.column width="50%"}
**What We've Accomplished:**
- Replaced direct $\psi$ learning with hierarchical embeddings
- Used attention mechanism to relate diseases to signatures
- Created contextualized representations for each disease-signature pair
- Maintained model-based approach while adding interpretability
:::

::: {.column width="50%"}
**Next Steps:**
- Implement in our existing model (clust_gp)
- Experiment with different embedding dimensions
- Add medication effects
- Validate on our dataset
:::

::::

::: {.callout-tip}
## For Your Real Application
With D=350, K=20, use L=5-8 for optimal parameter efficiency while maintaining interpretability.
:::
